Embracing the Laws of Physics: \\ Three Reversible Models of Computation

### 2018-06-15 survey/survey.tex
A further paragraph on the wires intuition. Expad on where Groupoid fits in. Citations for Rig Categories. Small edits.

\newcommand{\fin}[1]{\ensuremath{\left[#1\right]}}
\newcommand{\Nat}{\ensuremath{\mathbb{N}}}
\newcommand{\true}{\mathit{true}}
\newcommand{\false}{\mathit{false}}

@@

@@

In the previous section, we saw how we could take the proof
terms of commutative semirings as our starting point. What we need
now is to understand how proofs of algebraic identities should be
considered equivalent. Classical algebra does not help, as proofs
in algebra are not considered first-class citizens. However,
another route is available to us: since the work of 
Hofmann and Streicher~\cite{hofmann96thegroupoid}, we know that
one can model types as \emph{groupoids}.  Previously, we
modeled types as (finite) sets.

Thus, rather than looking at (untyped) commutative semirings,
we should looked at a \emph{typed} version. This process frequently
goes by the moniker of ``categorification.''  We want a categorical
algebra, where the basic objects are groupoids (to model our types),
and where there is a natural notion of $+$ and $*$.  At first,
we hit what seems like a serious stumbling block: the category of
all groupoids, Groupoid, does not have co-products or products.
We don't want to work internally in Groupoid -- we want operations
\emph{on} groupoids. In other words, we want something akin to
symmetric monoidal categories, but with two interacting 
monoidal structures.  Luckily, this already exists: the categorical
analog to (commutative) semirings are (symmetric) Rig Categories.
This straightforwardly generalizes to symmetric Rig Groupoids.

How does this help? Coherence conditions! Symmetric monoidal categories,
to start somewhere simple, do not just introduce weak transformations
like the associator $\alpha$ and the left $\lambda$ and right $\rho$
unitors, but also coherence conditions that these must satisfy. Looking,
for example, at just the additive fragment of $\Pi$ (i.e. with just $0$,
$1$ and $+$ for the types, $\odot$ and $\oplus$ as combinators, and
only the terms so expressible), the sub-language would correspond,
denotationally, to exactly symmetric monoidal groupoids. And what
these possess are exactly some \emph{equations between equations}
as commutative diagrams.  Adding in the coherence conditions 
corresponding to naturality of the various transformers gives us a
list of equations between programs.  Furthermore, the natural transformations
that arise are in fact natural isomorphisms -- all of which are
reversible.

>>>>

\newcommand{\fin}[1]{\ensuremath{\left[#1\right]}}
\newcommand{\Nat}{\ensuremath{\mathbb{N}}}
\newcommand{\true}{\mathit{true}}
\newcommand{\false}{\mathit{false}}
\newcommand{\Gpd}{\ensuremath{\mathsf{Groupoid}}}

@@

While we will not make the details of the stretchable wires and
slidable boxes formal, it is useful for intuition.  One caveat
though: some of the sliding and stretching needs to be done in
spaces of higher dimension than 3 to have ``enough room'' to
move things along without collision or over-stretching wires.
That, unfortunately, means that some equivalences are harder to
grasp. Luckily, most equivalences only need 3 dimensions.

@@

Previously we saw how we could take the proof terms of commutative semirings as
our starting point for $\Pi$. What we need
now is to understand how \emph{proofs} of algebraic identities should be
considered equivalent. Classical algebra does not help, as proofs
are not considered first-class citizens. However,
another route is available to us: since the work of 
Hofmann and Streicher~\cite{hofmann96thegroupoid}, we know that
one can model types as \emph{groupoids}.  The additional
structure comes from explicitly modeling the ``identity
types'': instead of regarding all terms which witness
the equality of (say) $a$ and $b$ of type $A$ as being
indistinguishible, we posit that there may in fact by many.
This alone is enough to show that types can be modeled by
groupoids.

Thus, rather than looking at (untyped) commutative semirings,
we should look at a \emph{typed} version. This process frequently
goes by the moniker of ``categorification.''  We want a categorical
algebra, where the basic objects are groupoids (to model our types),
and where there is a natural notion of $+$ and $*$.  At first,
we hit what seems like a serious stumbling block: the category of
all groupoids, \Gpd, does not have either co-products 
nor products. However, we don't want to work internally in
\Gpd -- we want operations
\emph{on} groupoids. In other words, we want something akin to
symmetric monoidal categories, but with two interacting 
monoidal structures.  Luckily, this already exists: the categorical
analog to (commutative) semirings are (symmetric) Rig
Categories~\cite{laplaza72,kelly74}.
This straightforwardly generalizes to symmetric Rig Groupoids.

How does this help? Coherence conditions! Symmetric monoidal categories,
to start somewhere simple, do not just introduce transformations
like the associator $\alpha$ and the left $\lambda$ and right $\rho$
unitors, but also coherence conditions that these must satisfy. Looking,
for example, at just the additive fragment of $\Pi$ (i.e. with just $0$,
$1$ and $+$ for the types, $\odot$ and $\oplus$ as combinators, and
only the terms so expressible), the sub-language would correspond,
denotationally, to exactly (non-empty) symmetric monoidal groupoids. And what
these possess are exactly some \emph{equations between equations}
as commutative diagrams.  Adding in the coherence conditions 
corresponding to naturality of the various transformers gives us a
list of equations between programs.  Furthermore, all the natural transformations
that arise are in fact natural isomorphisms -- all of which are
reversible.
